---
title: "Global Gender Inequality Index: 2010-2019"
author: "Jess Ebner-Statt"
date: "2024-10-26"
output: 
  html_document:
    theme:
      bootswatch: minty
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction and Data Sources
This document joins global spatial data with UN Human Development Data, creating a map of changes in the HDR Gender Inequality Index between 2010 and 2019.

Data sources are listed below for reproducibility purposes:

* **Global country shapefile**: World Countries Generalized geojson from [ArcGIS Hub](https://hub.arcgis.com/datasets/2b93b06dc0dc4e809d3c8db5cb96ba69_0/explore?location=-0.249646%2C0.000000%2C1.41).
* **Gender Inequality Index**: Excel file from the UNDP Human Development Reports' [Documentation and Downloads page](https://hdr.undp.org/data-center/documentation-and-downloads).
  * Filters used:
    * **Index:** GII
    * **Indicator:** Gender Inequality Index (value)
    * **Year:** 2010, 2019
    * **Group/Region/Country:** no filter applied
  * This was converted to a CSV locally, but no other formatting was changed before cleaning in R.
  
### Setting up R Environment
```{r Setting Up, message=FALSE, warning=FALSE, results='hide'}
#Load libraries
library(tidyverse)
library(countrycode)
library(sf)
library(janitor)
library(here)

#Load in data
shapefile <- st_read(here("homework4", "world_countries.geojson")) %>%
  clean_names(.) #WGS84

gii <- read_csv(here("homework4", "hdr-data.csv")) %>%
  clean_names()
```
### Inspecting and Cleaning GII Data
```{r Cleaning }
# Consider variable types
gii_data_types <- gii %>%
  summarise_all(class) %>%
  pivot_longer(everything(),
    names_to = "variable",
    values_to = "variable_type")
# All looks sensible
rm(gii_data_types)

# Drop redundant columns
gii_clean <- gii %>%
  dplyr::select(-index_code, -index, -dimension, -indicator_code,
                -indicator, -note)

# Remove any ISO codes beginning with "ZZ"; these are statistics for whole continents, HDI levels, or the globe
# (ChatGPT helped with the str_starts function)
gii_clean <- gii_clean %>%
  filter(., !str_starts(country_iso_code, "ZZ"))

# Pivot the table to allow easy comparison of 2010 and 2019 values
gii_pivoted <- gii_clean %>%
  pivot_wider(., names_from = year, values_from = value, names_prefix = "gii_")
# Count distinct countries in original dataset, to make sure this has been done properly
gii %>%
  filter(., !str_starts(country_iso_code, "ZZ")) %>%
  distinct(country_iso_code) %>%
  nrow() 
#165 - this matches. Note that many countries are missing
```
*n.b. Removing the aggregate statistics was not technically necessary, as these are going to be lost when joined to the spatial data anyway. Nevertheless, I removed them to create a "cleaner" GII dataset which can be analysed independently.*

### Calculate the Difference Metric
```{r Calculating Difference}
# Some countries have null values for one year
# Difference can therefore not be calculated for these, but I have kept them in the code anyway.
# Whoever is mapping can choose to change this if they wish.
gii_final <- gii_pivoted %>%
  mutate(diff = gii_2019 - gii_2010)
head(gii_final, 20)
```
Note that negative numbers in the `diff` column constitute a *decrease* in gender inequality over time.

### Joining GII data to the Shapefile
```{r Joining}
# shapefile has 2 digit ISOs
# gii_final has 3 digit ISOs

# Create a 2-digit ISO column in the gii_final df 
gii_final <- gii_final %>%
  mutate(iso2 = countrycode(country_iso_code, origin = "iso3c", destination = "iso2c"))

# Left join GII values onto shapefile (thereby preserving all countries in the final map)
shapefile_joined <- shapefile %>%
  left_join(., gii_final, by = c("iso" = "iso2"))%>%
  select(country.x, iso, gii_2010, gii_2019, diff, geometry)%>%
  rename(country = country.x)
```
*n.b. The countrycode package could also be used to connect countries to their parent countries (e.g. Puerto Rico -> US), but I'm not sure whether this is needed in the context of this homework.*

### Mapping
```{r Plot}
#Leaving this space empty for the person who is going to be creating the map as part of next week's homework
```